Autoregressive Connections in Probabilistic Circuits

### 2024-03-05 aaai2024_neural_circuits/neural_circuits_rebuttal.tex

%File: anonymous-submission-latex-2024.tex
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai24}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet}  % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS
%
% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
\usepackage{algorithm}
\usepackage{algorithmic}

%
% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
\usepackage{newfloat}
\usepackage{listings}
\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
\lstset{%
	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
	aboveskip=0pt,belowskip=0pt,%
	showstringspaces=false,tabsize=2,breaklines=true}
\floatstyle{ruled}
\newfloat{listing}{tb}{lst}{}
\floatname{listing}{Listing}
%
% Keep the \pdfinfo as shown here. There's no need
% for you to add the /Title and /Author tags.
\pdfinfo{
/TemplateVersion (2024.1)
}

% DISALLOWED PACKAGES
% \usepackage{authblk} -- This package is specifically forbidden
% \usepackage{balance} -- This package is specifically forbidden
% \usepackage{color (if used in text)
% \usepackage{CJK} -- This package is specifically forbidden
% \usepackage{float} -- This package is specifically forbidden
% \usepackage{flushend} -- This package is specifically forbidden
% \usepackage{fontenc} -- This package is specifically forbidden
% \usepackage{fullpage} -- This package is specifically forbidden
% \usepackage{geometry} -- This package is specifically forbidden
% \usepackage{grffile} -- This package is specifically forbidden
% \usepackage{hyperref} -- This package is specifically forbidden
% \usepackage{navigator} -- This package is specifically forbidden
% (or any other package that embeds links such as navigator or hyperref)
% \indentfirst} -- This package is specifically forbidden
% \layout} -- This package is specifically forbidden
% \multicol} -- This package is specifically forbidden
% \nameref} -- This package is specifically forbidden
% \usepackage{savetrees} -- This package is specifically forbidden
% \usepackage{setspace} -- This package is specifically forbidden
% \usepackage{stfloats} -- This package is specifically forbidden
% \usepackage{tabu} -- This package is specifically forbidden
% \usepackage{titlesec} -- This package is specifically forbidden
% \usepackage{tocbibind} -- This package is specifically forbidden
% \usepackage{ulem} -- This package is specifically forbidden
% \usepackage{wrapfig} -- This package is specifically forbidden
% DISALLOWED COMMANDS
% \nocopyright -- Your paper will not be published if you use this command
% \addtolength -- This command may not be used
% \balance -- This command may not be used
% \baselinestretch -- Your paper will not be published if you use this command
% \clearpage -- No page breaks of any kind may be used for the final version of your paper
% \columnsep -- This command may not be used
% \newpage -- No page breaks of any kind may be used for the final version of your paper
% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
% \pagestyle -- This command may not be used
% \tiny -- This is not an acceptable font size.
% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference

\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.

% The file aaai24.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%

% Title

\title{}

%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
% \author {
%     Pedro Zuidberg Dos Martires
% }
% \affiliations{
%     Örebro University
% }

\author{}
\affiliations{}

\usepackage{mathtools} % amsmath with fixes and additions
% \usepackage{siunitx} % for proper typesetting of numbers and units
\usepackage{booktabs} % commands to create good-looking tables
\usepackage{tikz}
\usetikzlibrary{decorations.pathreplacing,calligraphy,calc,hobby,intersections,through}
\usetikzlibrary{spn}

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fontsm
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.

\usepackage{multirow}
\usepackage{courier}
\usepackage{listings, lstautogobble,amsfonts}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{mdframed}
\usepackage{mathtools}
\usepackage{xspace}
\usepackage{xcolor}
\usepackage{caption}
\usepackage{multicol}
\usepackage{thmtools}
\usepackage{bm}
\usepackage{thm-restate}
\usepackage{todonotes}
\usepackage[inline]{enumitem}
\usepackage{soul}
\usepackage{physics}

% Recommended, but optional, packages for figures and better typesetting:
\usepackage{microtype}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{booktabs} % for professional tables
\usepackage{cancel}

\pgfplotsset{compat=1.18}

\input{macros}

%Commands definitions
\newcommand{\setbackgroundcolour}{\pagecolor[rgb]{0.19,0.19,0.19}}  
\newcommand{\settextcolour}{\color[rgb]{0.77,0.77,0.77}}    
\newcommand{\invertbackgroundtext}{\setbackgroundcolour\settextcolour}

%Command execution. 
%If this line is commented, then the appearance remains as usual.
% \invertbackgroundtext

\newcommand{\spacing}{\vspace{0.1cm}}
\newcommand{\question}[1]{\noindent{\bf\color{teal}{#1}}}

\newcommand{\answer}[1]{\noindent{\normalfont\color{black}{#1}}}

\begin{document}

% \maketitle
\noindent
We would like to thank the reviewers for their considerate and thoughtful comments. In order to adequately address these comments, we intend to purchase an extra page.

\vspace{0.1cm}
\noindent {\bf Reviewer 1}

\noindent We address the comment made by Reviewer 1 concerning marginalization in our responses to Reviewers 2 and 3.
\spacing
\question{
How do PNCs empirically compare to state-of-the-art (intractable) density estimators?
}

\answer{
State of the art neural density estimators usually deploy a plethora of tricks, such as the use of \textit{discretized logistic mixtures}~\Citep{salimans2017pixelcnn} to achieve a high performance. While we plan in future work to explore these ideas for PNCs, the main focus of the experiments was to show that giving up to a certain degree on tractability leads to more expressive models. Indeed, our experiments confirm this hypothesis, especially when comparing PNCs. PQCs, and state-of-the-art circuits.

Nevertheless, we also compare to less tractable models. In particular normalizing flows (IDF) and hierarchical latent variable models (BitSwap) (both using neural networks). In this regard we believe we make a fair effort in terms of comparisons. Especially when considering the parameter count of the models: The parameter count of PNCs is half the one of SHCLT and only a fraction of neural density estimators such as PixelCNN ($\approx56M$ parameters).

% In future work we plan on scaling up our approach (usually not trivial for PCs as parameter learning is tricky in massively overparametrized circuits). We would then compare to larger models, \eg PixelCNN++ ($\approx56M$ parameters). We would also deploy common tricks from the density estimation literature such as  in the leaves \Citep{salimans2017pixelcnn}.
}

\vspace{0.1cm}
\noindent {\bf Reviewer 2}

\noindent
\question{Relationship between CPCs and PNCs.}

\answer{
The advantage of introducing CPCs lies with the neat interpretation as deep mixture of Bayesian networks and the fact that we can, for instance, proof their validity (Proposition 3.9) using arguments similar to the case of PCs.

We agree, however, with the reviewer that simply stating the functional form of sum units for PNCs (3.10 Definition) obscures to a certain degree their relationship to CPCs. For the camera-ready version we will include an extended version of the following discussion, which gives us PNCs starting from CPCs.

Consider a sum unit of a CPC:
\begin{talign}
	\sum_{j\in\inputs(k)} \weight_{kj} \circuit_j(\xvars_n  {\mid}  \xvars_{\parents(n)})
\end{talign}
This can be rearranged into:
\begin{talign}
	\sum_{j\in\inputs(k)}
		\weight_{kj}
		\frac{
			\circuit_j(  \xvars_{\parents(n)} \mid \xvars_n  )
		}{
			\circuit_j(  \xvars_{\parents(n)  })
		}
	\circuit_j(  \xvars_n)
\end{talign}
We approximate the ratio with a neural network as follows:
\begin{talign}
	\nweight_{kj}(\xvars_{\ancestors(n)})\approx 		\weight_{kj}
	\frac{
		\circuit_j(  \xvars_{\parents(n)} \mid \xvars_n  )
	}{
		\circuit_j(  \xvars_{\parents(n)  })
	}.
\end{talign}
This gives us the expression of Definition 3.10. Constraining the $ \nweight_{kj} ( \xvars_{\ancestors(n)})$ to sum to one together with necessitating $\circuit_j (\xvars_n)$ being a valid circuit then gives us the validity of the PNC. This lets us conclude that PNCs are neural approximations of CPCs. We will make this explicit in the final version of the paper.
Note that it is this approximation that makes PNCs tractable: PNCs only need a single circuits for each $j$ while CPCs need a circuit for every $j$ and every instantiation of $\xvars_{\parents(n)}$.
}

\spacing
\question{
The marginal inference algorithm was a bit unclear. Does it require m number of passes through the circuit, or can it be done in a single pass? [...]
}

The marginalization method we describe in Proposition 3.13 walks through a circuit in a depth-first fashion. This requires only visiting each node at most once (assuming proper caching). This can indeed be translated into a breadth-first circuit traversal, which is then easily parallelized for layered PNCs. This would then only require a single pass through the layered computation graph.
As pointed out by the reviewer this would indeed boil down to replacing distributions at the leaves with $1$ and evaluating the circuit.
We will add a discussion on this to the paper as well as provide pseudocode.

% This is what we refer to as "the delicate point" in the paper (after the proof of Proposition 3.13).

\spacing
\question{
Also, I think PNCs are tractable for more general ordered conditionals than shown in Corollary 3.14. [...].
}

Yes, indeed thank you for the suggestion. We will add it to the paper. Showing this is quite straightforward by combining Proposition 3.13 (marginalizing $X_m$) and Corollary 3.14.

\spacing
\question{Minor Questions / Comments}

\answer{
Due to space constraints, we are, unfortunately, not able to comment on them here. We will address them in the camera-ready version of the manuscript.
}

% \begin{itemize}

% \item Definition 3.2: $pa(n)$ where $X_n$ is a set of variables was not previously defined. Is it just the union of parents of variables in $X_n$ (minus $X_n$ itself)?
% \item Eq (5): missing parenthesis for $w_{kj}$. Also, should the sum unit be $p_k(x_n | x_{an(n)})$ instead because the weights depend on $x_{an(n)}$, and should $p_j$ also be conditioned the same way for the recursive definition?
% \item Typo: Theorem 3.9 sum of $w_m$ -> $w_{nm}$
% \item Typo: Figure 5 caption “while on the left” -> “while on the right”
% \item Section 6.3: I did not follow “two components in the leaf units instead of 256”
% \item Are partition trees related to vtrees? (Kisa et al. "Probabilistic sentential decision diagrams." 2014)
% \item Definition 3.10 seems to assume Boolean variables. Can this be applied to continuous variables?

% \end{itemize}

\vspace{0.1cm}
\noindent {\bf Reviewer 3}

\noindent
Similarly to Reviewer 1, Reviewer 3 commented on tractability of marginals (in practice). In the final version of the manuscript we will include a more detailed discussion. Here we already answer some of the questions.

	\question{For the PNC structures used for 28x28 images, have the following questions:
	}

	\question{
	A. What is the missingness pattern that is allowed in the images so we still get tractable queries? Is there a intuitive way to see which missing patterns are allowed for the pixels?
	}
	\answer{Yes, there is an intuitive pattern. Consider the partition tree in Figure 3. We can then marginalize out the variables the right branch of the tree. For images this is similar but with a second dimension.}

	\question{
	 B. Is the missingness pattern allowed for MARs the same for PNCs learned on the different datasets? Or is it data dependent (assume still images of same size).
	}
	\answer{The missingness pattern is the same for different datasets (assuming the same PNC structure is used.)}

	\question{
	 C. In practice how often we get actual tractability for MARs for images.
	}
	\answer{In practice one could, for instance, marginalize the right side of an image and have a model for the left side. This means that one would need to know beforehand the queries posed to the model. An interesting question is whether this can be mitigated using approximate inference techniques such as sampling or adapting a form of loopy belief propagation.
	Alternatively, one could focus on tasks where only ordered marginalization (PNC) is needed instead of any-order marginalization (PC). As we suggest in our future work section, arithmetic coding seems to be a good candidate and there might be other such tasks of interest. 
	}

\bibliography{references}

\end{document}

### 2024-07-01 202409_nesy_neural_circuits/main.tex
all

%File: anonymous-submission-latex-2024.tex
\documentclass[letterpaper]{article} % DO NOT CHANGE THIS
\usepackage{aaai24}  % DO NOT CHANGE THIS
\usepackage{times}  % DO NOT CHANGE THIS
\usepackage{helvet}  % DO NOT CHANGE THIS
\usepackage{courier}  % DO NOT CHANGE THIS
\usepackage[hyphens]{url}  % DO NOT CHANGE THIS
\usepackage{graphicx} % DO NOT CHANGE THIS
\urlstyle{rm} % DO NOT CHANGE THIS
\def\UrlFont{\rm}  % DO NOT CHANGE THIS
\usepackage{natbib}  % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\usepackage{caption} % DO NOT CHANGE THIS AND DO NOT ADD ANY OPTIONS TO IT
\frenchspacing  % DO NOT CHANGE THIS
\setlength{\pdfpagewidth}{8.5in} % DO NOT CHANGE THIS
\setlength{\pdfpageheight}{11in} % DO NOT CHANGE THIS
%
% These are recommended to typeset algorithms but not required. See the subsubsection on algorithms. Remove them if you don't have algorithms in your paper.
\usepackage{algorithm}
\usepackage{algpseudocode}

%
% These are are recommended to typeset listings but not required. See the subsubsection on listing. Remove this block if you don't have listings in your paper.
\usepackage{newfloat}
\usepackage{listings}
\DeclareCaptionStyle{ruled}{labelfont=normalfont,labelsep=colon,strut=off} % DO NOT CHANGE THIS
\lstset{%
	basicstyle={\footnotesize\ttfamily},% footnotesize acceptable for monospace
	numbers=left,numberstyle=\footnotesize,xleftmargin=2em,% show line numbers, remove this entire line if you don't want the numbers.
	aboveskip=0pt,belowskip=0pt,%
	showstringspaces=false,tabsize=2,breaklines=true}
\floatstyle{ruled}
\newfloat{listing}{tb}{lst}{}
\floatname{listing}{Listing}
%
% Keep the \pdfinfo as shown here. There's no need
% for you to add the /Title and /Author tags.
\pdfinfo{
/TemplateVersion (2024.1)
}

% DISALLOWED PACKAGES
% \usepackage{authblk} -- This package is specifically forbidden
% \usepackage{balance} -- This package is specifically forbidden
% \usepackage{color (if used in text)
% \usepackage{CJK} -- This package is specifically forbidden
% \usepackage{float} -- This package is specifically forbidden
% \usepackage{flushend} -- This package is specifically forbidden
% \usepackage{fontenc} -- This package is specifically forbidden
% \usepackage{fullpage} -- This package is specifically forbidden
% \usepackage{geometry} -- This package is specifically forbidden
% \usepackage{grffile} -- This package is specifically forbidden
% \usepackage{hyperref} -- This package is specifically forbidden
% \usepackage{navigator} -- This package is specifically forbidden
% (or any other package that embeds links such as navigator or hyperref)
% \indentfirst} -- This package is specifically forbidden
% \layout} -- This package is specifically forbidden
% \multicol} -- This package is specifically forbidden
% \nameref} -- This package is specifically forbidden
% \usepackage{savetrees} -- This package is specifically forbidden
% \usepackage{setspace} -- This package is specifically forbidden
% \usepackage{stfloats} -- This package is specifically forbidden
% \usepackage{tabu} -- This package is specifically forbidden
% \usepackage{titlesec} -- This package is specifically forbidden
% \usepackage{tocbibind} -- This package is specifically forbidden
% \usepackage{ulem} -- This package is specifically forbidden
% \usepackage{wrapfig} -- This package is specifically forbidden
% DISALLOWED COMMANDS
% \nocopyright -- Your paper will not be published if you use this command
% \addtolength -- This command may not be used
% \balance -- This command may not be used
% \baselinestretch -- Your paper will not be published if you use this command
% \clearpage -- No page breaks of any kind may be used for the final version of your paper
% \columnsep -- This command may not be used
% \newpage -- No page breaks of any kind may be used for the final version of your paper
% \pagebreak -- No page breaks of any kind may be used for the final version of your paperr
% \pagestyle -- This command may not be used
% \tiny -- This is not an acceptable font size.
% \vspace{- -- No negative value may be used in proximity of a caption, figure, table, section, subsection, subsubsection, or reference
% \vskip{- -- No negative value may be used to alter spacing above or below a caption, figure, table, section, subsection, subsubsection, or reference

\setcounter{secnumdepth}{2} %May be changed to 1 or 2 if section numbers are desired.

% The file aaai24.sty is the style file for AAAI Press
% proceedings, working notes, and technical reports.
%

% Title

\title{Probabilistic Neural Circuits}

%Example, Single Author, ->> remove \iffalse,\fi and place them surrounding AAAI title to use it
\author {
    Pedro Zuidberg Dos Martires
}
\affiliations{
    Centre for Applied Autonomous Sensor Systems (AASS), Örebro University, Sweden
}

% \author{}
% \affiliations{}

@@

\begin{document}

\maketitle

\begin{abstract}
    Probabilistic circuits (PCs) have gained prominence in recent years as a versatile framework for discussing probabilistic models that support tractable queries and are yet expressive enough to model complex probability distributions.
    Nevertheless, tractability comes at a cost: PCs are less expressive than neural networks.
    In this paper we introduce probabilistic neural circuits (PNCs), which strike a balance between PCs and neural nets in terms of tractability and expressive power. Theoretically, we show that PNCs can be interpreted as deep mixtures of Bayesian networks. Experimentally, we demonstrate that PNCs constitute powerful function approximators.
    % In this paper we explore This paper explores the interplay between PCs and neural networks, presenting the concept of Probabilistic Neural Circuits (PNCs) as an intermediary model. PNCs are introduced as deep mixtures of Bayesian networks, offering a structured approach to capturing complex dependencies in data.
\end{abstract}

@@

In recent years probabilistic circuits (PCs) (also called sum-product networks)~\citep{darwiche2003differential,poon2011sum} have emerged as an assembly language to talk about tractable probabilistic models and inference therein~\citep{vergari2021compositional}. The core idea is quite simple: we start with a set of independent random variables and construct complex probability distribution by recursively adding and multiplying them together.
There are two common ways of interpreting PCs. Firstly, we can consider them to be hierarchical mixture models. Secondly, we look at them as neural nets consisting of sums, products, and atomic probability distributions.

Most of the recent advances in the field adhere to the second perspective: use an overparametrized probabilistic model and fit it to data using gradient based methods by leveraging discrete GPUs~\citep{peharz2019random,dang2021juice}. The computation units of such circuits are organized in a layered fashion. We give an example in Figure~\ref{fig:circuit}.

\begin{figure}[t]
    \centering
    \resizebox{0.85\columnwidth}{!}{
        \tikzset{point/.style={circle,inner sep=0pt,minimum size=3pt,fill=red}}

@@

    \end{tikzpicture}
    }
    \caption{Layered probabilistic circuit following the construction of~\citep{shih2021hyperspns}. Data (modeled as random variables) is first fed into the leaf layer at the bottom. The output of the leaf layer is a mixture of distributions produced by the sum units. In the sum-product layer (Layer 2)  mixtures of random variables are combined by taking pairwise products, these are then again mixed using sum units. Finally, the root layer (at the top) gives us the joint probability distribution. The red edges indicate functional dependencies not present in traditional probabilistic circuits but present in probabilistic neural circuits.}
    \label{fig:circuit}
\end{figure}

A major advantage of PCs is their ability to answer certain queries in polynomial time -- given that adequate restrictions are imposed on a circuit's structure~\citep{vergari2021compositional}. An example of such a tractable query would be the computation of conditional probabilities for so-called \textit{smooth and decomposable} PCs~\citep{darwiche2001decomposable,darwiche2003differential}.

This tractability, however, comes at a hefty price: the properties imposed on PCs in order to ensure polynomial time queries limit their expressive power~\citep{martens2014expressive,sharir2018sum,zhang2021probabilistic}.
This is in contrast to general neural networks and even sum-product networks with fewer structural constraints~\citep{delalleau2011shallow,kileel2019expressive}. \citet{martens2014expressive} have shown that decomposability is in fact a necessary condition for tractable marginal inference.

Nevertheless, using the concepts of \textit{conditional smoothness} and \textit{conditional decomposability}~\citep{sharir2018sum}, we study in this paper the space of models that lie in between probabilistic circuits and neural networks.
Concretely, we make the following contributions:
\begin{enumerate}
    \item We introduce conditional probabilistic circuits, from which we construct  probabilistic neural circuits (PNCs), which we interpret as deep mixtures of Bayesian nets.
    \item  We provide a prescription to construct layered PNCs.
    \item We provide an implementation of layered PNCs and experimentally study their expressive power.
\end{enumerate}

Our work is influenced by that of \citet{sharir2018sum}. We discuss the relationship to their approach (dubbed sum-product-quotient networks) in Section~\ref{sec:related}.

\section{Preliminaries}
\label{sec:prelim}

In the remainder of the paper we will denote random variables by uppercase $\Xvar$'s, the corresponding values are lowercase $\xvar$'s. Sets of random variables and their corresponding values are typed in boldface: $\Xvars$ and $\xvars$, respectively.
The definitions and notions we introduce in this section are loosely based on the work of~\citet{vergari2021compositional}.

\begin{definition}[Probabilistic Circuit]
    \label{def:circuit}
    A probabilistic circuit over random variables $\Xvars$ is a parametrized computational graph encoding a probability density function $\circuit(\Xvars{=}\xvars)$.
    The circuit consists of three kinds of computational units:
    \textit{leaf}, \textit{product}, and \textit{sum}.
    Each product or sum unit receives inputs from a set of input units denoted by $\inputs(n)$.
    Each unit $k$ encodes a function $\circuit_{k}(\cdot)$ as follows:
    \begin{align*}
        {\circuit}_k(\xvars_n)=
        \begin{cases}
            f_k({\xvars_n})                                             & \text{if $k$ leaf unit}    \\
            \circuit_{k_l}(\xvars_{n_l})  \circuit_{k_r} (\xvars_{n_r}) & \text{if $k$ product unit} \\
            \sum_{j\in\inputs(k)} \weight_{kj} \circuit_j(\xvars_n)     & \text{if $k$ sum unit}
        \end{cases}
        % \label{eq:def:circuit}
    \end{align*}
    where $f_k(\xvars_n)$  denotes a parametrized probability distribution having as support the sample space of the random variables in $\Xvars_n$.
\end{definition}

\begin{definition}[Scope]
    \label{def:scope:cond}
    The scope of a unit $k$, denoted by $\scope(k)$, is the set of random variables $\Xvars_n$ for which the function $\circuit_k (\cdot)$ encodes a probability distribution.
\end{definition}

Two important properties that are usually imposed on probabilistic circuits are smoothness and decomposability as they allow for tractable queries, \eg computing marginals~\citep{darwiche2001decomposable,darwiche2003differential}.
\begin{definition}[Smoothness]
    A circuit is smooth if for every sum unit $k$ its inputs encode distributions over the same random variables:
    $\forall j_1, j_2 {\in} \inputs(k)$ it holds that $\scope(j_1){=}\scope(j_2) $.
\end{definition}

\begin{definition}[Decomposability]
    A circuit is decomposable if the inputs of every product unit
    $k$ encode distributions over disjoint sets of random variables:
    $\scope (k_l) \cap \scope (k_r) = \emptyset$ with $\{k_l, k_r\}= \inputs (k)$.
\end{definition}

\begin{definition}[Valid Probabilsitic Circuit]
    We call a probabilistic circuit valid if for every unit $k$ we have that $p_k(\xvars_n){\geq} 0$ and $\int p_k(\xvars_n) \differential \xvars_n{=}1$.\footnote{Note that our notion of validity is slightly stricter than in its original definition, \cf \citep{poon2011sum}}
\end{definition}

As discussed by \citet{peharz2015theoretical}, probabilistic circuits are valid if they are smooth, decomposable, and that for the weights in the sum units we have $\weight_{kj} \in \mathbb{R}^+$ and $\sum_{j\in\inputs(k)} \weight_{kj} = 1$ for every $k$. Note that the notation in Definition~\ref{def:circuit} already suggests that the circuit is smooth as the inputs to the sum units are functions over the same set of variables $\Xvars_n$.

Furthermore, we can assume, without loss of generality, that sum and product units occur in an alternating fashion in the circuit~\citep{peharz2020einsum}.
This observation naturally leads to a layer-wise construction of probabilistic circuits where consecutive layers alternate between sum and product layers.
Such layered probabilistic circuits~\citep{peharz2019random} have the advantage that the computations within a layer can be trivially parallelized.
We can further abstract the layers in a circuit by fusing together alternating sums and products into a single sum-product layer~\citep{peharz2020einsum}.

In Figure~\ref{fig:circuit} we give a graphical representation of a layered circuit. Layers consist of  blocks of computational units that are processed sequentially in a bottom-up fashion. Layers are themselves constituted of so-called partitions. The circuit in Figure~\ref{fig:circuit} has four partitions in the leaf layer, two in the sum-product layer, and a single partition in the root layer. By construction, partitions in the same layer have disjoint scopes. Moreover, partitions are further subdivided into input components and output components, which constitute the elemental computing units. The circuit in Figure~\ref{fig:circuit} has, except at the very bottom and top, three such input and output components in each partition.

We can uniquely identify each computational unit (or component) in the circuit by specifying the layer, the partition, whether it is an input or an output, and its position within a partition. Counting layers from bottom to top, and partitions and units from left to right. Each component can be identified using 4 indices: $\component_{l,p,i,c}$. The first index $l$ identifies the layer, the second $p$ the partition, the third $i \in \{1,2 \}$ whether it is an input or output, and the fourth $c$ the horizontal position within a partition.
For instance, the symbol $\component_{2223}$ corresponds to  the upper-right unit in the sum-product layer.

\section{Conditional Probabilistic Circuits}
\label{sec:cpc}

We will first introduce the notion of posets (partially ordered sets) of random variables (Section~\ref{sec:poset}).
This will allow us to generalize probabilistic circuits to conditional probabilistic circuits, which we interpret as deep mixtures of Bayesian networks (Section ~\ref{sec:dmbn}).
We then introduce probabilistic neural circuits and their tractable queries (Section~\ref{sec:pnc}).

\subsection{Partially Ordered Random Variables}
\label{sec:poset}

Consider a set of random variables $\Xvars {=}\{\Xvar_1,\dots, \Xvar_N \}$ on which we impose the \textit{parents} relationship $\parents(\cdot)$.
The parents relationship induces a directed acyclic graph on the random variables $\Xvars$, where the nodes are the random variables themselves and an edge is present between $\Xvar_i$ and $\Xvar_j$ if $\Xvar_i {\in} \parents(\Xvar_j)$.
This gives us a partial ordering of the variables $\Xvars$.
We also define the ancestor relationship $\ancestors(\cdot)$ to be the transitive closure of $\parents(\cdot)$. That is, the ancestors of a random variable are its direct parents and recursively their parents.
Furthermore, we denote the poset for the random variables $\Xvars$ by $\mathcal{O}(\Xvars)$. We say that the relationship $\Xvars_i {\sqsubset} \Xvars_j$ between two sets holds if
$
    \forall \Xvar_r {\in} \Xvars_i, \Xvar_q {\in} \Xvars_j : \Xvar_q {\notin}  \ancestors(\Xvars_r)
$.
We also define the relation $\Xvar_i\sqsubset \Xvar_j$ on random variables as $\{\Xvar_i \}{\sqsubset} \{\Xvar_j \}$.

\begin{example}[Bayesian Network]
    Partially ordered random variables induce a factorization of a joint probability distribution. A prominent example of such a factorization are Bayesian networks (\cf Figure ~\ref{fig:bn}) where we have:
    \begin{talign}
        \label{eq:bn}
        p(\Xvars{=}\xvars) = \prod_{n:\Xvar_n \in \Xvars} p_n(X_n{=}x_n \mid \Xvars_{\parents(n)}{=}\xvars_{\parents(n)}).
    \end{talign}
\end{example}

\begin{figure}

    \begin{minipage}{0.50\linewidth}
        \begin{align*}
            X_1 \sqsubset X_2 &  & X_1 \sqsubset X_4 \\
            X_1 \sqsubset X_3 &  & X_3 \sqsubset X_1 \\
            X_3 \sqsubset X_2 &  & X_3 \sqsubset X_4 \\
            X_2 \sqsubset X_3 &  & X_2 \sqsubset X_4 \\
        \end{align*}

    \end{minipage}%
    \begin{minipage}{0.49\linewidth}
        \resizebox{0.9\linewidth}{!}{

            \centering

            \begin{tikzpicture}
                % Nodes
                \node[draw, circle,ultra thick] (x3) {$X_3$};
                \node[draw, circle, below=of x3,ultra thick] (x2) {$X_2$};
                \node[draw, circle,  right=2cm of $(x3)!0.5!(x2)$3,ultra thick] (x4) {$X_4$};
                \node[draw, circle, left=of x2,ultra thick] (x1) {$X_1$};

                % Edges
                \draw[->, ultra thick] (x1) -- (x2);
                \draw[->,ultra thick] (x2) -- (x4);
                \draw[->, ultra thick] (x3) -- (x4);
            \end{tikzpicture}
        }
    \end{minipage}
    \caption{Right: Bayesian network. Left: partial order relations that hold.}
    \label{fig:bn}
\end{figure}

In the example above we denote $\parents(X_n)$ by $\Xvars_{\parents(n)}$. This will allow us to omit the random variable $\Xvar_n$ when writing down a probability distribution and only use the instantiation $\xvar_n$ instead.

\subsection{Deep Mixtures of Bayesian Networks}
\label{sec:dmbn}

>>>>

%% The first command in your LaTeX source must be the \documentclass command.
%%
%% Options:
%% twocolumn : Two column layout.
%% hf: enable header and footer.
\documentclass[
	% twocolumn,
	% hf,
]{ceurart}

%%
%% One can fix some overfulls
\sloppy

%%
%% Minted listings support 
%% Need pygment <http://pygments.org/> <http://pypi.python.org/pypi/Pygments>
\usepackage{listings}
%% auto break lines
\lstset{breaklines=true}

@@

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% Rights management information.
%% CC-BY is default license.
\copyrightyear{2022}
\copyrightclause{Copyright for this paper by its authors.
	Use permitted under Creative Commons License Attribution 4.0
	International (CC BY 4.0).}

%%
%% This command is for the conference information
\conference{18th International conference on
	Neural-Symbolic Learning and Reasoning}

%%
%% The "title" command
\title{Autoregressive Connections in Probabilistic Circuits}

%%
%% The "author" command and its associated commands are used to define
%% the authors and their affiliations.
\author{Pedro {Zuidberg Dos Martires}}
\address{Örebro University, Sweden}

%%
%% The abstract is a short summary of the work to be presented in the
%% article.
\begin{abstract}
	Probabilistic circuits (PCs) have gained prominence in recent years as a versatile framework for discussing probabilistic models that support tractable queries and are yet expressive enough to model complex probability distributions.
	Nevertheless, tractability comes at a cost: PCs are less expressive than neural networks.
	In this extended abstract we briefly summarize the concept of a probabilistic neural circuit (PNC), which strikes a balance between PCs and neural nets in terms of tractability and expressive power, and distinguishes itself from ordinary PCs by the presence of autoregressive connections.
\end{abstract}

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\begin{keywords}
	LaTeX class \sep
	paper template \sep
	paper formatting \sep
	CEUR-WS
\end{keywords}

%%
%% This command processes the author and affiliation and title
%% information and builds the first part of the formatted document.
\maketitle

@@

In recent years probabilistic circuits (PCs) (also called sum-product networks)~\citep{darwiche2003differential} have emerged as an assembly language to talk about tractable probabilistic models and inference therein~\citep{vergari2021compositional}. The core idea is quite simple: we start with a set of independent random variables and construct complex probability distribution by recursively adding and multiplying them together.
There are two common ways of interpreting PCs. Firstly, we can consider them to be hierarchical mixture models. Secondly, we look at them as neural nets consisting of sums, products, and atomic probability distributions.

A major advantage of PCs is their ability to answer certain queries in polynomial time -- given that adequate restrictions are imposed on a circuit's structure~\citep{vergari2021compositional}. An example of such a tractable query would be the computation of conditional probabilities for so-called \textit{smooth and decomposable} PCs~\citep{darwiche2001decomposable,darwiche2003differential}.
This tractability, however, comes at a hefty price: the properties imposed on PCs in order to ensure polynomial time queries limit their expressive power~\citep{martens2014expressive,sharir2018sum}.
In this extended abstract we give a short exposition of the ideas presented in~\citep{zuidberg2024probabilistic} where we used the concepts of \textit{conditional smoothness} and \textit{conditional decomposability}~\citep{sharir2018sum}
to develop probabilistic neural circuits -- a novel probabilistic model class that trades-off tractability and expressive power.

The key difference between (layered) probabilistic circuits ~\citep{peharz2020einsum} and probabilistic neural circuits is the presence of áutoregressive connections in the circuit. We illustrate this in Figure~\ref{fig:partition_graph}. It is precisely these autoregressive connections that lead to a huge boost in expressive power. We show this in our experimental study in~\citep{zuidberg2024probabilistic}. Note that this gain in expressive power comes at the cost of tractability: not all marginals of the joint distributions are computable in poly-time anymore.

\begin{figure}[t]
	\begin{minipage}[c]{0.49\linewidth}
		\centering
		\resizebox{0.85\columnwidth}{!}{
			\tikzset{point/.style={circle,inner sep=0pt,minimum size=3pt,fill=red}}

@@

		\end{tikzpicture}
		}
	\end{minipage}
	\begin{minipage}[c]{0.49\linewidth}
		\centering
		\resizebox{0.9\columnwidth}{!}{
			\begin{tikzpicture}
